\textbf{COMP0085 Summative Assignment}

### 2023-01-04 submission.tex
add ard for loop bp and edit report

\documentclass[12pt]{article}
\usepackage{graphicx}
\usepackage[margin=2cm]{geometry}
\usepackage[utf8]{inputenc}
\usepackage{tikz}
\usepackage[export]{adjustbox}
\usepackage{indentfirst}
\usepackage{wrapfig}
\usepackage{listings}
\usepackage{color}
\usepackage{enumerate}
\usepackage{amssymb, bm}
\usepackage{amsmath}
\usepackage{csvsimple}
\usepackage{tikz}
\usetikzlibrary{positioning}
\usepackage{lscape}
\newcommand\tab[1][1cm]{\hspace*{#1}}
\definecolor{mygreen}{rgb}{0,0.6,0}
\definecolor{mygray}{rgb}{0.5,0.5,0.5}
\definecolor{mymauve}{rgb}{0.58,0,0.82}
\lstset{ %
    backgroundcolor=\color{gray!10!white},
  basicstyle=\tiny, %footnotesize,        % the size of the fonts that are used for the code
  breakatwhitespace=false,         % sets if automatic breaks should only happen at whitespace
  breaklines=true,                 % sets automatic line breaking
  captionpos=b,                    % sets the caption-position to bottom
  commentstyle=\color{mygreen},    % comment style
  deletekeywords={...},            % if you want to delete keywords from the given language
  escapeinside={\%*}{*)},          % if you want to add LaTeX within your code
  extendedchars=true,              % lets you use non-ASCII characters; for 8-bits encodings only, does not work with UTF-8
  frame=single,	                   % adds a frame around the code
  keepspaces=true,                 % keeps spaces in text, useful for keeping indentation of code (possibly needs columns=flexible)
  keywordstyle=\color{blue},       % keyword style
  language=Python,                 % the language of the code
  morekeywords={*,...},           % if you want to add more keywords to the set
  numbers=left,                    % where to put the line-numbers; possible values are (none, left, right)
  numbersep=10pt,                   % how far the line-numbers are from the code
  numberstyle=\tiny\color{mygray}, % the style that is used for the line-numbers
  rulecolor=\color{black},         % if not set, the frame-color may be changed on line-breaks within not-black text (e.g. comments (green here))
  showspaces=false,                % show spaces everywhere adding particular underscores; it overrides 'showstringspaces'
  showstringspaces=false,          % underline spaces within strings only
  showtabs=false,                  % show tabs within strings adding particular underscores
  stepnumber=1,                    % the step between two line-numbers. If it's 1, each line will be numbered
  stringstyle=\color{mymauve},     % string literal style
  tabsize=1,	                   % sets default tabsize to 2 spaces
  title=\lstname                   % show the filename of files included with \lstinputlisting; also try caption instead of title
}
\graphicspath{ }
\usetikzlibrary{arrows}

\title{\textbf{COMP0085 Summative Assignment}}
%\author{Jian Shu (James) Wu \\ }
\date{Jan 4, 2023}

@@

The set $\{A, D, E, F, H\}$ is a non-unique smallest set of molecules such that if the concentrations of the species within the set are known, the concentrations of the others $\{B, C, G, I\}$ would all be independent (conditioned on the measured ones).
\newpage

@@

Thus, we would expect to recover the factors of $A$ and $C$, $\delta[A]$ and $\delta[C]$, the two parent nodes of the observations.
\newpage

@@

From factor analysis, we would recover $\delta[A]$ and $\delta[C]$.
In theory, if these concentration perturbations had unknown parent factors, we could use the results to recover the concentration perturbations of those parent features.
However, in our DAG this is not hte case, so we would not be able to recover any other species in the cascade because  they are all downstream from the known concentration perturbations and those recovered from factor analysis.

@@

\[k(s, t) = \theta^2 \left( \exp\left( - \frac{2\sin^2(\pi(s-t)/\tau)}{\sigma^2}\right) + \phi^2 \exp\left( - \frac{(s-t)^2}{2\eta^2}\right)\right) + \zeta^2 \delta_{s=t}\]

@@

\item [$\sigma$:] \begin{itemize}
\item The dot product $(\textbf{x}^{T} \textbf{x})$ where $\textbf{x} \in \mathbb{R}^{N \times D}$ is  $\mathcal{O}(D^2N)$
\item The dot product $\mu^{T} \mu$ where $\mu \in \mathbb{R}^{D \times K}$ is  $\mathcal{O}(K^2D)$
\item The dot product $(\mu^{T} \mu)\text{ESS}$ where $\mu^{T} \mu \in \mathbb{R}^{K \times K}$ and $\text{ESS} \in \mathbb{R}^{K \times K}$ is $\mathcal{O}(K^3)$
\item The dot product  $\text{ESS}^T\textbf{x}$ where $\text{ES} \in \mathbb{R}^{N \times K}$ and $\textbf{x} \in \mathbb{R}^{N \times D}$ is  $\mathcal{O}(KND)$
\item The dot product $(\text{ESS}^T\textbf{x}) \mu$ where $\text{ESS}^T\textbf{x} \in \mathbb{R}^{K\times D}$ and $\mu \in \mathbb{R}^{D \times K}$ is $\mathcal{O}(K^2 D)$
\end{itemize}
\item [$\pi$:] \begin{itemize}
    \item The mean operation for $\text{ES} \in \mathbb{R}^{N \times K}$ along the first dimension is $\mathcal{O}(NK)$
    \end{itemize}
\end{itemize}

@@

\[\textbf{x} = \sum_{k=1}^K\pi_k s_k + \epsilon \text{ and } \sum_{k=1}^K s_k = 1\]

where $\epsilon \sim \mathcal{N}(0, \Sigma_{\epsilon})$. This also wouldn't be appropriate because it assumes a dirichlet distribution on the vector $\textbf{s}$. Our data is constructed by having each $s_k$ independently sampled from a Bernoulli distribution for that feature.

@@

where $\epsilon \sim \mathcal{N}(0, \sigma^2\textbf{I})$ and $P(\textbf{s}) = \prod_{k=1}^K P(s_k)$. This is appropriate for our data because $P(s_k)$ is our Bernoulli distribution and we are linearly combining different features with $\textbf{w} in \mathbb{R}^{D \times K}$and then adding noise $\epsilon$.

@@

Similar to question 3, we can use automatic relevance determination (ARD) as a hyperparameter method to select relevant features by placing a prior on $\mu_i$. With a hyper-M step, certain features will have diverging precision, indicating that they are not relevant to the model output. Thus, the number of remaining features will be our selection for $K$.

@@

\newpage
\section*{Appendix 1: abstract\_binary\_latent\_factor\_model.py}
\lstinputlisting[language=Python]{src/models/binary_latent_factor_models/abstract_binary_latent_factor_model.py}

\newpage
\section*{Appendix 2: abstract\_binary\_latent\_factor\_approximation.py}
\lstinputlisting[language=Python]{src/models/binary_latent_factor_approximations/abstract_binary_latent_factor_approximation.py}

\newpage
\section*{Appendix 3: main.py}
\lstinputlisting[language=Python]{main.py}

\newpage
\section*{Appendix 4: constants.py}
\lstinputlisting[language=Python]{src/constants.py}

\newpage
\section*{Appendix 5: generate\_images.py}
\lstinputlisting[language=Python]{src/generate_images.py}
\newpage
\section*{Appendix 6: MStep.py}
\lstinputlisting[language=Python]{demo_code/MStep.py}
\end{document}

>>>>

\documentclass[12pt]{article}
\usepackage{graphicx}
\usepackage[margin=2cm]{geometry}
\usepackage[utf8]{inputenc}
\usepackage{tikz}
\usepackage[export]{adjustbox}
\usepackage{indentfirst}
\usepackage{wrapfig}
\usepackage{listings}
\usepackage{color}
\usepackage{enumerate}
\usepackage{amssymb, bm}
\usepackage{amsmath}
\usepackage{csvsimple}
\usepackage{tikz}
\usetikzlibrary{positioning}
\usepackage{lscape}
\newcommand\tab[1][1cm]{\hspace*{#1}}
\definecolor{mygreen}{rgb}{0,0.6,0}
\definecolor{mygray}{rgb}{0.5,0.5,0.5}
\definecolor{mymauve}{rgb}{0.58,0,0.82}
\lstset{ %
    backgroundcolor=\color{gray!10!white},
  basicstyle=\tiny, %footnotesize,        % the size of the fonts that are used for the code
  breakatwhitespace=false,         % sets if automatic breaks should only happen at whitespace
  breaklines=true,                 % sets automatic line breaking
  captionpos=b,                    % sets the caption-position to bottom
  commentstyle=\color{mygreen},    % comment style
  deletekeywords={...},            % if you want to delete keywords from the given language
  escapeinside={\%*}{*)},          % if you want to add LaTeX within your code
  extendedchars=true,              % lets you use non-ASCII characters; for 8-bits encodings only, does not work with UTF-8
  frame=single,	                   % adds a frame around the code
  keepspaces=true,                 % keeps spaces in text, useful for keeping indentation of code (possibly needs columns=flexible)
  keywordstyle=\color{blue},       % keyword style
  language=Python,                 % the language of the code
  morekeywords={*,...},           % if you want to add more keywords to the set
  numbers=left,                    % where to put the line-numbers; possible values are (none, left, right)
  numbersep=10pt,                   % how far the line-numbers are from the code
  numberstyle=\tiny\color{mygray}, % the style that is used for the line-numbers
  rulecolor=\color{black},         % if not set, the frame-color may be changed on line-breaks within not-black text (e.g. comments (green here))
  showspaces=false,                % show spaces everywhere adding particular underscores; it overrides 'showstringspaces'
  showstringspaces=false,          % underline spaces within strings only
  showtabs=false,                  % show tabs within strings adding particular underscores
  stepnumber=1,                    % the step between two line-numbers. If it's 1, each line will be numbered
  stringstyle=\color{mymauve},     % string literal style
  tabsize=1,	                   % sets default tabsize to 2 spaces
  title=\lstname                   % show the filename of files included with \lstinputlisting; also try caption instead of title
}
\graphicspath{ }
\usetikzlibrary{arrows}
\usepackage{titlesec}

\titleformat{\section}
  {\normalfont\fontsize{14}{15}\bfseries}{\thesection}{1em}{}
\titleformat{\subsection}
  {\normalfont\fontsize{12}{15}\bfseries}{\thesection}{1em}{}

\title{\textbf{COMP0085 Summative Assignment}}
%\author{Jian Shu (James) Wu \\ }
\date{Jan 6, 2023}

@@

The set $\{A, D, E, F, H\}$ is a non-unique smallest set of molecules such that if the concentrations of the species within the set are known, the concentrations of the others $\{B, C, G, I\}$ would all be independent (conditioned on the measured ones).

@@

Thus, we would expect to recover the factors of $A$ and $C$, $\delta[A]$ and $\delta[C]$, the two parent nodes of the observations.

@@

From factor analysis, we would recover $\delta[A]$ and $\delta[C]$.
In theory, if these concentration perturbations had unknown parent factors, we could use the results to recover the concentration perturbations of those parent features.
However, in our DAG this is not hte case, so we would not be able to recover any other species in the cascade because  they are all downstream from the known concentration perturbations and those recovered from factor analysis.
Similarly, the downstream weights in the DAG from the known measurements would also be unidentifiable.
Using factor analysis, we would only be able to identify the upstream weights from the measurements, up to an unknown scale factor.
These are the weights $\Lambda_{BA}$, $\Lambda_{DC}$, and $\Lambda_{EC}$.
Moreover, knowing $\delta[C]$ from factor analysis, we would also be able to determine $\Lambda_{CA}$, up to an unknown scale factor.

@@

\[k(s, t) = \theta^2 \exp\left( - \frac{2\sin^2(\pi(s-t)/\tau)}{\sigma^2}\right) + \phi^2 \exp\left( - \frac{(s-t)^2}{2\eta^2}\right) + \zeta^2 \delta_{s=t}\]

@@

\item [$\sigma$:] \begin{itemize}
\item The dot product $(\textbf{x}^{T} \textbf{x})$ where $\textbf{x} \in \mathbb{R}^{N \times D}$ is  $\mathcal{O}(D^2N)$
\item The dot product $\mu^{T} \mu$ where $\mu \in \mathbb{R}^{D \times K}$ is  $\mathcal{O}(K^2D)$
\item The dot product $(\mu^{T} \mu)\text{ESS}$ where $\mu^{T} \mu \in \mathbb{R}^{K \times K}$ and $\text{ESS} \in \mathbb{R}^{K \times K}$ is $\mathcal{O}(K^3)$
\item The dot product  $\text{ES}^T\textbf{x}$ where $\text{ES} \in \mathbb{R}^{N \times K}$ and $\textbf{x} \in \mathbb{R}^{N \times D}$ is  $\mathcal{O}(KND)$
\item The dot product $(\text{ES}^T\textbf{x}) \mu$ where $\text{ES}^T\textbf{x} \in \mathbb{R}^{K\times D}$ and $\mu \in \mathbb{R}^{D \times K}$ is $\mathcal{O}(K^2 D)$
\end{itemize}
\item [$\pi$:] \begin{itemize}
    \item The mean operation for $\text{ES} \in \mathbb{R}^{N \times K}$ along the first dimension is $\mathcal{O}(NK)$
    \end{itemize}
\end{itemize}

@@

\[\textbf{x} = \sum_{k=1}^K s_k \mu_k + \epsilon \text{ and } \sum_{k=1}^K s_k = 1\]

where $\epsilon \sim \mathcal{N}(0, \Sigma_{\epsilon})$. This also wouldn't be appropriate because it assumes a dirichlet distribution on the vector $\textbf{s}$. Our data is constructed by having each $s_k$ independently sampled from a Bernoulli distribution indicating the presence of feature $\mu_k$.

@@

where $\epsilon \sim \mathcal{N}(0, \sigma^2\textbf{I})$ and $P(\textbf{s}) = \prod_{k=1}^K P(s_k)$. This is appropriate for our data because $P(s_k)$ is our Bernoulli distribution and we are linearly combining different features with $\textbf{W} \in \mathbb{R}^{D \times K}$and then adding noise $\epsilon$.

@@

\newpage
\subsection*{(d)}

We can use automatic relevance determination (ARD) as a hyperparameter method to select relevant features by placing a prior on $\mu_i$ in the same way as Question 4. With a hyper-M step, certain features will have diverging precision, leaving us with the relevant features, from which we get our $K$. We can implement this by running the message passing algorithm (loopy BP) which was implemented for Question 6 for the expectation step and the VB algorithm for the maximisation step (along with its hyper-M Step):

\begin{figure}[h]
\centering
\includegraphics[scale=0.4]{outputs/q5/d-latent-factors-comparison}
\caption{Learned Latent Factors vs Inverse Alpha}
\label{fig:}
\end{figure}

We can see that ARD has more difficulty finding relevant features when using loop BP.
Looking at the features themselves, we can see that they are not very clear compared to the mean field approximation or Variational Bayes in Question 4.
This is probably why we do not consistently have the same number of $\alpha_k$ values that don't diverge as compared with Question 3.
However, we can see for $K<8$ that all of the features are deemed relevant as we would expect.

\newpage
Comparing the free energy plots of models trained on different $K$ values:

\begin{figure}[h]
\centering
\includegraphics[scale=0.4]{outputs/q5/d-free-energy}
\caption{Free Energy for different values of k}
\label{fig:}
\end{figure}

We can see the difficulties of this approach due to the inability for loopy BP to converge for this particular problem.
This likely contributes to the difficulty for ARD to find relevant features when the message passing algorithm is unable to converge on generating meaningful features.
To improve this, we could also put priors on the other parameters $\sigma$ and $\pi$ in the hopes that this will help stabilise the algorithm.
Moreover, because loopy BP is unable to converge for this problem, it is quite computationally intensive as EM will always run the maximum number of iterations without improvement in the results.
Knowing that it won't converge, employing early stopping or fewer iterations can also be helpful.

\newpage
The Python code for question 5:
\lstinputlisting[language=Python]{src/solutions/q5.py}

@@

\newpage
\section*{Appendix 1: abstract\_binary\_latent\_factor\_model.py}
\lstinputlisting[language=Python]{src/models/binary_latent_factor_models/abstract_binary_latent_factor_model.py}
\newpage
\section*{Appendix 2: abstract\_binary\_latent\_factor\_approximation.py}
\lstinputlisting[language=Python]{src/models/binary_latent_factor_approximations/abstract_binary_latent_factor_approximation.py}
\newpage
\section*{Appendix 3: main.py}
\lstinputlisting[language=Python]{main.py}
\newpage
\section*{Appendix 4: constants.py}
\lstinputlisting[language=Python]{src/constants.py}
\newpage
\section*{Appendix 5: generate\_images.py}
\lstinputlisting[language=Python]{src/generate_images.py}
\newpage
\section*{Appendix 6: m\_step.py}
\lstinputlisting[language=Python]{demo_code/m_step.py}
\end{document}
